Fabric-Elasticity Relationships of Tibial Trabecular Bone are Similar in Osteogenesis Imperfecta and Healthy Individuals

### 2021-07-22 09_Article/Simon_2021_Bone_Draft_V03.tex
Article corrections

\subsection{Participants}
The healthy group included a total of 120 patients from a previous reproducibility study performed at the University Department of osteoporosis in Bern \cite{Schenk2020}. The group was composed of 64 females and 56 males aged between 20 and 92 years old with a mean age of 32 $\pm$ 15 years. These subjects had not taken any medication known to affect bone metabolism nor presented with any prior osteoporosis fracture. The second group was scanned as part of the ASTEROID study (NCT number: NCT03118570) at different locations, namely at the Shriners Hospital for Children-Canada. The study coordination was done by the McGill University in Montreal. This group was composed of 35 females and 15 males with confirmed diagnosis of OI Type I, III or IV. There were 35 patients diagnosed with OI type I, 2 with type III, and 13 with type IV. The participants of the OI group were aged between 19 and 69 years old with a mean age of 44 $\pm$ 14 years.

\subsection{HR-pQCT}
HR-pQCT scans (XtremeCTII, SCANCO Medical, Br\"{u}tisellen,
Switzerland) were performed at the distal tibia on all patients from both groups. Participants in the healthy group were scanned using an in-house protocol as described in \cite{Schenk2020}. Namely, the reference line was positioned at the proximal margin of the dense structure formed by the tibia plafond and three stacks were scanned proximal to this reference line, see Figure \ref{01_Healthy}. On the other hand, participants in the OI group were scanned using the manufacturer's standard protocol i.e. the reference line is placed at the subchondral endplate of the ankle joint and one stack was scanned at 22.5 mm proximal to the reference line (standard clinical section) \cite{Whittier2020}, see Figure \ref{01_OI}. The scanned region according to the two different protocol are shown side by side in Figure \ref{01_ClinicalSections}. For both the healthy and the OI groups, each stack consisted of 168 slices and an isotropic voxel resolution of 61 \si{\micro}m. This led to a thickness of roughly 10.2 mm for each stack. Standardized scanning settings were used (voltage of 60 kVp, 900 $\mu$A, 100 ms integration time) for the healthy group as well as for the OI group. For the healthy group, motion artefacts of first, middle and last slice (i.e slices number 1, 252, and 504) were graded on a scale proposed by \citeauthor{Pialat2012} \cite{Pialat2012} of 1 (no motion artefacts) to 5 (extreme motion artefacts). The final grade of each scan was defined as the highest slice grade. For the OI group, as the scan consisted of one stack, only one grade was attributed using the same scale as for the control group. Scans were then processed independently of their quality grading. A summary of the scans grading is shown in Figure \ref{01_MotionArtefacts}.

@@

In general, six cubic ROIs were selected at random position in each scan in a predefined area. Each ROI had to contain trabecular bone, but no cortical bone. For the healthy group, the ROIs were selected in the most proximal stack uniquely (see Figure \ref{01_Healthy} stack number 3) to be at rougthly the same anatomical location as for the OI group. Then for both the healthy group and the OI group, the stack was divided into two halves and the ROIs were selected to have the centers of three ROIs in the proximal half and three in the distal half, see Figure \ref{01_ROISelection}. For one individual diagnosed with OI type III it was not possible to extract any ROI as there was no enough trabecular bone. This led to 720 healthy ROIs from 120 individuals and 294 OI ROIs from 49 individuals.\\

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]
	{Pictures/01_ROISelection_Example}
	\caption{Middle slices example of ROIs selection for HR-pQCT reconstruction of a healthy tibia. Cortical bone is in white and trabecular bone in gray. The stacks are separated by red lines and cyan squares represent the ROIs. The stack used for analysis is higlighted in cyan and the dashed rose line represents the middle of the stack.}
	\label{01_ROISelection}
\end{figure}

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth, trim= 0 0 0 100]
	{Pictures/01_FabricExample}
	\caption{Typical ROI with the visualization of its fabric tensor using MIL method. Eigenvectors of the fabric tensor define its orientation and eigenvalues set lengths of the ellipsoid radii. DA is the ratio between the highest and the lowest eigenvalue.}
	\label{01_FabricExample}
\end{figure}

A \si{\micro}FE analysis was performed using \textsc{ABAQUS 6.14}. In brief, each voxel of the cleaned ROI was converted to a fully integrated linear brick elements (C3D8) using a direct voxel conversion approach. Then, a stiffness $E$ of 10,000 MPa and a Poisson's ratio $\nu$ of 0.3 were assigned to each element. The homogenisation process consisted of 6 independent simulations of different load cases, 3 uni-axial and 3 simple shear cases, using kinematic uniform boundary conditions (KUBCs) \cite{Panyasantisuk2015}. Unlike periodicity-compatible mixed uniform boundary conditions (PMUBCs), KUBCs do not require to rotate the ROI into the respective fabric coordinate system. Such rotation would potentially decrease image quality. The homogenization process allowed to calculate the components of the stiffness tensor and to calibrate the parameters of the Zysset-Curnier fabric-elasticity model \cite{Zysset1995}. This model compute the fourth order stiffness tensor $\mathbb{S}$ using the BV/TV or $\rho$, fabric tensor $\mathbf{M}$, three elasticity parameters $\lambda_0$, $\lambda_0$', and $\mu_0$, and two exponents, $k$ and $l$, as shown in Equation \ref{Eq202}.\\

@@

\subsubsection{Statistical Significance}
The general significance level was set to 95\% for all tests. Confidence intervals were computed for t-tested variables to quantify the difference in both groups means. As Mann-Whitney tests are performed on the median, only the corresponding p-value is presented. Finally, non-parametric permutation tests are less powerful but give an empirical 95\% exclusion range and a p-value. If the difference in means belong to this exclusion range, it can be stated that group means are different with 95\% certainty.

@@

The linear regression was performed using the \textsc{statsmodels} package from \textsc{Python 3.6}. Linear regression quality for the TMD analysis was assessed using the Pearson correlation coefficient ($R^2$) and the standard error of the estimate (SE). Linear regression on Zysset-Curnier model was assessed using the adjusted Pearson correlation coefficient squared ($R^2_{adj}$) and relative error between the orthotropic observed and the predicted tensor using norm of fourth-order tensors (NE), see Equation \ref{Eq205} and \ref{Eq206}.

@@

\subsection{Linear Regression with Original Data Sets}
Figure \ref{02_GeneralRegression} shows the results of the linear regression analysis of each group separately, between the values of the observed stiffness tensors from \si{\micro}FE simulations and the predicted values using the Zysset-Curnier model\cite{Zysset1995} and the parameters obtained after performing the linear regression with linear mixed-effect model. The dashed line represents the 1:1 correlation. For the healthy group (Figure \ref{02_Healthy}) the fit is performed on 720 ROIs leading to 8640 data points. The $R^2_{adj}$ is slightly above 0.95 and the NE is of 18\% $\pm$ 10\%. The linear regression analysis for the OI group (Figure \ref{02_OI}) performed on 294 ROIs led to 3528 data points, an $R^2_{adj}$ close to 0.85 and a NE of 62\% $\pm$ 233\%. It can be noticed that, as the values of the observed stiffness tensor decreases, data points tend to be further apart from the diagonal (dashed line). Moreover, data the points from the stiffness tensors with lowest values are exclusively above the diagonal. The range of stiffness tensors of the OI group is wider compared to the one of the healthy group and ROIs with lower BV/TV present lower stiffness values. The values of these ROIs stiffness tensors components tend to be overestimated by the fit.\\

@@

As the previous studies \cite{Folkestad2012,Kocijan2015,Rolvien2018} have the same age range as our matched groups, we can compare the morphological parameters. The CT imaging system explains most of the differences between the absolute morphological values of the present study compared to the others. \citeauthor{Folkestad2012}\cite{Folkestad2012}, \citeauthor{Kocijan2015}\cite{Kocijan2015}, and \citeauthor{Rolvien2018}\cite{Rolvien2018} have performed their measurements on first generation XCT scanners with a voxel size of 82 \si{\micro}m, while we have used a second generation XCT with a voxel size of 61 \si{\micro}m. The work from \citeauthor{Agarwal2016}\cite{Agarwal2016} investigated differences between the two scanner types. They showed that BV/TV, Tb.Th., and Tb.Sp. are higher in second generation XCT scanners and Tb.N. is lower compared to first generation XCT. These findings are in full agreement with our observations. Another potential bias is introduced by the fact that the present study analyses the median values of six cubic ROIs with 5.3 mm side length. This conditions the Tb.N. and Tb.Sp. as they depend on the ROI size. Moreover, conditions imposed for random ROI selection can lead to further biased values, especially for OI patients, as the ROI must contain at least some portion of trabecular bone. Even with the low sample size (2x28 individuals), the statistical tests have shown significant differences between groups, with the CV presenting the most significant difference. The CV shows that heterogeneity of OI trabecular bone is higher compared to healthy control and is consequently the more discriminant parameter. Finally, the significant differences observed in BV/TV and DA, even with matched age and sex, justifies the choice of a variable matching for fabric-elasticity relationships analysis, because the linear regression must be performed on similar ranges to obtain comparable values. \\

@@

The analysis of TMD gives interesting outputs as well. The t-test reveal a higher TMD in OI trabecular bone than in healthy condition with 95\% certainty and a very high significance level, even with similar BV/TV and DA. The linear regression slope CI being strictly different from zero shows that a relation exists between TMD and BV/TV. This can have different origins. From a biological point of view, a mineralisation gradient develops between the core of the trabecula undergoing secondary mineralisation and the outer surface subjected to remodelling and deposition of younger, less mineralised matrix. As trabecular thickness decreases with BV/TV, this means that a lower BV/TV could lead to a lower mineralization of the core of the trabeculae. The second explanation for this slope comes from the scanning. Effectively, during scanning a phenomenon called partial volume effect occurs and its impact decreases with an increasing BV/TV as the ratio of bone surface to bone volume BS/BV decreases. The former (biological) explanation is expected to have a lower impact than the latter (scanning). Regarding the other coefficients of the linear regression, the CI of group variable excludes zero as well, meaning that the intercept depends on the group. This could be explained by the bisphosphonate treatment that OI patients receive. Effectively, bisphosphonates are aimed to slow down the remodelling process which lead to higher (secondary) mineralization of the bone. According to the results of \citeauthor{Indermaur2021}\cite{Indermaur2021} these findings suggest adding a correction accounting for the TMD in FE simulations to catch the higher modulus, ultimate stress and post-yield behaviour of OI bone compared to healthy bone at the ECM level. Regarding the interaction between BV/TV and the group, the p-value obtained showed a high non-significance meaning that the slopes of both OI and healthy groups are the same. This could be visualised in Figure \ref{02_TMD} where the fitted lines are obtained using the fixed effects of the model and the group variable if fixed. In this plot, performed with the interaction term (BV/TV x Group) the fitted lines appear to be quasi-parallel. \\
The main limitations of this study are the definition of "homogeneous enough" (i.e. low CV) and the fact that it is limited to  XCTII reconstructions of the distal tibia. Moreover, having only one patient with OI type III where we could extract ROIs does not allows to do statistics. %As those patients are less active, it could be interesting to analyse the impact of this condition on the weight bearing tibia. 

@@

A similar study could be performed on radius XCTII reconstructions to generalise the above findings. Another limitation is that the measurements were performed on different HR-pQCT devices which were not cross-calibrated. However, as they are the same model (XCTII) and the quality control measurement ensures $\pm$ 1\% precision on BMD, this is expected to have a minor impact. \\

>>>>

\subsection{Participants}
The healthy group included a total of 120 patients from a previous reproducibility study performed at the University Department of osteoporosis in Bern \cite{Schenk2020}. The group was composed of 64 females and 56 males aged between 20 and 92 years old with a mean age of 32 $\pm$ 15 years. These subjects had not taken any medication known to affect bone metabolism nor presented with any prior osteoporosis fracture. The second group was scanned as part of the ASTEROID study (NCT number: NCT03118570) at 6 HR-pQCT (XtremeCT II) imaging sites in North America. The study coordination was done by the team at Shriners Hospital for Children-Canada/McGill University in Montreal. This group was composed of 35 females and 15 males with confirmed diagnosis of OI Type I, III or IV. There were 35 patients diagnosed with OI type I, 2 with type III, and 13 with type IV. The participants of the OI group were aged between 19 and 69 years old with a mean age of 44 $\pm$ 14 years.

\subsection{HR-pQCT}
HR-pQCT imaging (XtremeCTII, SCANCO Medical, Br\"{u}tisellen,
Switzerland) was performed at the distal tibia on all patients from both groups. Participants in the healthy group were scanned using an in-house protocol as described in \cite{Schenk2020}. Namely, the reference line was positioned at the proximal margin of the dense structure formed by the tibia plafond and three stacks were scanned proximal to this reference line, see Figure \ref{01_Healthy}. On the other hand, for participants in the OI group, the reference line is placed at the subchondral endplate of the ankle joint.  The mid-point of the scanned volume of interest was located at a distance of 7\% of total tibia length from the reference line \cite{GHASEMZADEH2017,Bonaretti2017}. Then, the scans were acquired according to the manufacturer's standard in vivo acquisition protocol, see Figure \ref{01_OI}. The scanned region according to the two different protocol are shown side by side in Figure \ref{01_ClinicalSections}. For both the healthy and the OI groups, each stack consisted of 168 slices and an isotropic voxel resolution of 61 \si{\micro}m. This led to a thickness of roughly 10.2 mm for each stack. Standardized scanning settings were used (voltage of 60 kVp, 900 $\mu$A, 100 ms integration time) for the healthy group as well as for the OI group. For the healthy group, motion artefacts of first, middle and last slice (i.e slices number 1, 252, and 504) were graded on a scale proposed by \citeauthor{Pialat2012} \cite{Pialat2012} of 1 (no motion artefacts) to 5 (extreme motion artefacts). The final grade of each scan was defined as the highest slice grade. For the OI group, as the scan consisted of one stack, only one grade was attributed using the same scale as for the control group. Scans were then processed independently of their quality grading. A summary of the scans grading is shown in Figure \ref{01_MotionArtefacts}.

@@

In general, six cubic ROIs were selected at random position in each scan in a predefined area. Each ROI had to contain trabecular bone, but no cortical bone. For the healthy group, the ROIs were selected in the most proximal stack uniquely (see Figure \ref{01_Healthy} stack number 3) to be at roughly the same anatomical location as for the OI group. Then for both the healthy group and the OI group, the stack was divided into two halves and the ROIs were selected to have the centers of three ROIs in the proximal half and three in the distal half, see Figure \ref{01_ROISelection}. For one individual diagnosed with OI type III it was not possible to extract any ROI as there was no enough trabecular bone. This led to 720 healthy ROIs from 120 individuals and 294 OI ROIs from 49 individuals.\\

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]
	{Pictures/01_ROISelection_Example}
	\caption{Example of randomly selected cubic ROIs (cyan squares) on a healthy bone in two planar views (left and right). Different stacks are separated by red lines, the stack used for analysis is highlighted in cyan, and the middle slice of the stack is represented by the dashed rose line. Cortical bone is in white and trabecular bone in gray.}
	\label{01_ROISelection}
\end{figure}

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth, trim= 0 0 0 100]
	{Pictures/01_FabricExample}
	\caption{Typical ROI with the visualization of its fabric tensor (colored ellipsoid) using MIL method. Eigenvectors of the fabric tensor define its orientation and eigenvalues set lengths of the ellipsoid radii. DA is the ratio between the highest and the lowest eigenvalue.}
	\label{01_FabricExample}
\end{figure}

A \si{\micro}FE analysis was performed using \textsc{ABAQUS 6.14}. In brief, each voxel of the cleaned ROI was converted to a fully integrated linear brick elements (C3D8) using a direct voxel conversion approach. Then, a Young's modulus $E$ of 10,000 MPa and a Poisson's ratio $\nu$ of 0.3 were assigned to each element. The homogenisation process consisted of 6 independent simulations of different load cases, 3 uni-axial and 3 simple shear cases, using kinematic uniform boundary conditions (KUBCs) \cite{Panyasantisuk2015}. Unlike periodicity-compatible mixed uniform boundary conditions (PMUBCs), KUBCs do not require to rotate the ROI into the respective fabric coordinate system. Such rotation would potentially decrease image quality. The homogenization process allowed to calculate the components of the stiffness tensor and to calibrate the parameters of the Zysset-Curnier fabric-elasticity model \cite{Zysset1995}. This model computes the fourth order stiffness tensor $\mathbb{S}$ using the BV/TV or $\rho$, fabric tensor $\mathbf{M}$, three elasticity parameters $\lambda_0$, $\lambda_0$', and $\mu_0$, and two exponents, $k$ and $l$, as shown in Equation \ref{Eq202}.\\

@@

\subsubsection{Statistical Significance}
The general confidence level was set to 95\% for all tests. Confidence intervals were computed for t-tested variables to quantify the difference in both groups means. As Mann-Whitney tests are performed on the median, only the corresponding p-value is presented. Finally, non-parametric permutation tests are less powerful but give an empirical 95\% exclusion range and a p-value. If the difference in means belong to this exclusion range, it can be stated that group means are different with 95\% certainty.

@@

The linear regression was performed using the \textsc{statsmodels} package from \textsc{Python 3.6}. Linear regression quality for the TMD analysis was assessed using the Pearson correlation coefficient squared ($R^2$) and the standard error of the estimate (SE). Linear regression on Zysset-Curnier model was assessed using the adjusted Pearson correlation coefficient squared ($R^2_{adj}$) and relative error between the orthotropic observed and the predicted tensor using norm of fourth-order tensors (NE), see Equation \ref{Eq205} and \ref{Eq206}.

@@

\subsection{Linear Regression with Original Data Sets}
Figure \ref{02_GeneralRegression} shows the results of the linear regression analysis of each group separately, between the values of the observed stiffness tensors from \si{\micro}FE simulations and the predicted values using the Zysset-Curnier model\cite{Zysset1995} and the parameters obtained after performing the linear regression with linear mixed-effect model. The dashed line represents the 1:1 correlation. For the healthy group (Figure \ref{02_Healthy}) the fit is performed on 720 ROIs leading to 8640 data points. The $R^2_{adj}$ is slightly above 0.95 and the NE is of 18\% $\pm$ 10\%. The linear regression analysis for the OI group (Figure \ref{02_OI}) performed on 294 ROIs led to 3528 data points, an $R^2_{adj}$ close to 0.85 and a NE of 62\% $\pm$ 233\%. It can be noticed that, as the values of the observed stiffness tensor decreases, data points tend to be further apart from the diagonal (dashed line). Moreover, the data points from the stiffness tensors with lowest values are exclusively above the diagonal. The range of stiffness tensors of the OI group is wider compared to the one of the healthy group and ROIs with lower BV/TV present lower stiffness values. The values of these ROIs stiffness tensors components tend to be overestimated by the fit.\\

@@

As the previous studies \cite{Folkestad2012,Kocijan2015,Rolvien2018} have the same age range as our matched groups, we can compare the morphological parameters. The CT imaging system explains most of the differences between the absolute morphological values of the present study compared to the others. \citeauthor{Folkestad2012}\cite{Folkestad2012}, \citeauthor{Kocijan2015}\cite{Kocijan2015}, and \citeauthor{Rolvien2018}\cite{Rolvien2018} have performed their measurements on first generation XCT scanners with a voxel size of 82 \si{\micro}m, while we have used a second generation XCT with a voxel size of 61 \si{\micro}m. The work from \citeauthor{Agarwal2016}\cite{Agarwal2016} investigated differences between the two scanner types. They showed that BV/TV, Tb.Th., and Tb.Sp. are higher in second generation XCT scanners and Tb.N. is lower compared to first generation XCT. These findings are in full agreement with our observations. Another potential bias is introduced by the fact that the present study analyses the median values of six cubic ROIs with 5.3 mm side length. This influences the Tb.N. and Tb.Sp. as they depend on the ROI size. Moreover, conditions imposed for random ROI selection can lead to further biased values, especially for OI patients, as the ROI must contain at least some portion of trabecular bone. Even with the low sample size (2x28 individuals), the statistical tests have shown significant differences between groups, with the CV presenting the most significant difference. The CV shows that heterogeneity of OI trabecular bone is higher compared to healthy control and is consequently the more discriminant parameter. Finally, the significant differences observed in BV/TV and DA, even with matched age and sex, justifies the choice of a variable matching for fabric-elasticity relationships analysis, because the linear regression must be performed on similar ranges to obtain comparable values. \\

@@

The analysis of TMD gives interesting outputs as well. The t-test reveal a higher TMD in OI trabecular bone than in healthy condition, agreeing with \citeauthor{Boyde1999}\cite{Boyde1999}, with 95\% certainty and a very high significance level, even with similar BV/TV and DA. The linear regression slope CI being strictly different from zero shows that a relation exists between TMD and BV/TV. This can have different origins. From a biological point of view, a mineralisation gradient develops between the core of the trabecula undergoing secondary mineralisation and the outer surface subjected to remodelling and deposition of younger, less mineralised matrix. As trabecular thickness decreases with BV/TV, this means that a lower BV/TV could lead to a lower mineralization of the core of the trabeculae. The second explanation for this slope comes from the scanning. Effectively, during scanning a phenomenon called partial volume effect occurs and its impact decreases with an increasing BV/TV as the ratio of bone surface to bone volume BS/BV decreases. The former (biological) explanation is expected to have a lower impact than the latter (scanning). Regarding the other coefficients of the linear regression, the CI of group variable excludes zero as well, meaning that the intercept depends on the group. This could be explained by the bisphosphonate treatment that OI patients receive. Effectively, bisphosphonates are aimed to slow down the remodelling process which lead to higher (secondary) mineralization of the bone. According to the results of \citeauthor{Indermaur2021}\cite{Indermaur2021} these findings suggest adding a correction accounting for the TMD in FE simulations to catch the higher modulus, ultimate stress and post-yield behaviour of OI bone compared to healthy bone at the ECM level. Regarding the interaction between BV/TV and the group, the p-value obtained showed a high non-significance meaning that the slopes of both OI and healthy groups are the same. This could be visualised in Figure \ref{02_TMD} where the fitted lines are obtained using the fixed effects of the model and the group variable if fixed. In this plot, performed with the interaction term (BV/TV x Group) the fitted lines appear to be quasi-parallel. \\
The main limitations of this study are the definition of "homogeneous enough" (i.e. low CV) and the fact that it is limited to  XCTII reconstructions of the distal tibia. Moreover, having only one patient with OI type III where we could extract ROIs does not allows to perform statistics. %As those patients are less active, it could be interesting to analyse the impact of this condition on the weight bearing tibia. 

@@

A similar study could be performed on radius XCTII reconstructions to generalise the above findings. Another limitation is the anatomical site scanned in control and OI group. The section scanned for the control group was at a fixed distance. On the other hand, the section scanned for the OI group was placed at a distance relative to the total length of the tibia. Nevertheless, this relative distance is aimed to be at the same location that the standard clinical section on average. Moreover, this study investigates mainly the fabric-elasticity relationships of trabecular bone and ROIs were matched to compare only architectural differences between healthy and OI trabecular bone. A final limitation is that the measurements were performed on different HR-pQCT devices which were not cross-calibrated. However, as they are the same model (XCTII) and the quality control measurement ensures $\pm$ 1\% precision on BMD, this is expected to have a minor impact. \\
